Perturbing SU($n$)-symmetric interactions

### 2020-01-26 collective_sun/sun_perturbation.tex
minor change to language
simplified some math

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQFS \sum_O \sum_{p<q} w_{Opq} O_{0,1}
  \EQFS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-body
operator $O_{pq}$.  If the two-body operator $O=X\otimes Y$, then
\begin{align}
  H_2^{(1)} \EQFS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} - \col{XY}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V2_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \f12 \sum_{p\ne q} \sp{2\bar h_{pq} w_{Opq}
    + \sum_k \p{h_{pk} w_{Okq} + w_{Opk} h_{kq}}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_{pq} - \f12\p{h_p + h_q},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ (indexed by $\Delta$) is an eigenvector
of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w} = \Delta \v{\m w},
  \label{eq:cond_2}
\end{align}
where
\begin{align}
  \check{\m h} \equiv 2\sum_{p<q} \bar h_{pq} \op{\set{p,q}}
  + \sum_{p<q} \ket{\set{p,q}} \sum_{k\notin\set{p,q}}
  \p{h_{pk} \bra{\set{k,q}} + h_{kq} \bra{\set{p,k}}}.
  \label{eq:h_super_mat}
\end{align}
Solving the general eigenvalue problem in \eqref{eq:cond_2} requires
diagonalizing the ${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$
matrix $\check{\m h}$.  When both he interaction Hamiltonian $H_0$
translationally invariant, however, this eigenvalue problem can be
reduced to that of diagonalizing an $\sim N\times N$ matrix; we
discuss this reduction in Appendix \ref{sec:trans_inv}.

@@

We wish to simplify
\begin{align}
  H_0 \V_1 \ket\psi
  = \sum_{X,k} \sum_{p<q} h_{pq} v_{Xk} \Pi_{pq} X_k \ket\psi,
  \label{eq:H_V1_psi_start}
\end{align}
which has terms with $k\in\set{p,q}$, and terms with
$k\notin\set{p,q}$.  In the case of $k\notin\set{p,q}$, the
permutation operator $\Pi_{pq}$ commutes with $X_k$ and annihilates on
$\ket\psi$, leaving terms at each fixed $k$ of the form
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}} h_{pq}
  = \sum_{p<q} h_{pq} - \sum_{\substack{p<q\\k\in\set{p,q}}} h_{pq}
  = E_0 - \f12 \sum_{\substack{p,q\\k\in\set{p,q}}} h_{pq}
  = E_0 - h_k,
\end{align}
where we define $h_{pq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $h_{pq}=h_{qp}$ and $h_{pp}=0$; and
$h_k \equiv \sum_\ell h_{k\ell}$.  The terms in
\eqref{eq:H_V1_psi_start} with $k\notin\set{p,q}$ are then
\begin{align}
  \sum_X \sum_{\substack{p<q\\k\notin\set{p,q}}}
  h_{pq} v_{Xk} \Pi_{pq} X_k \ket\psi
  = \sum_{X,k} \p{E_0 - h_k} v_{Xk} X_k \ket\psi
  = E_0 \V_1 \ket\psi - \sum_{X,k} h_k v_{Xk} X_k \ket\psi
\end{align}
while the terms in \eqref{eq:H_V1_psi_start} with $k\in\set{p,q}$ for
each fixed $X$ take the form
\begin{align}
  \sum_{\substack{p<q\\k\in\set{p,q}}}
  h_{pq} v_{Xk} \Pi_{pq} X_k \ket\psi
  = \sum_{k<q} h_{kq} v_{Xk} X_q \ket\psi
  + \sum_{p<k} h_{pk} v_{Xk} X_p \ket\psi
  = \sum_{k,q} h_{kq} v_{Xk} X_q \ket\psi
\end{align}
which implies
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \sum_X \p{\sum_{p,q} h_{pq} v_{Xp} X_q - \sum_p h_p v_{Xp} X_p}
  \ket\psi.
\end{align}

@@

We wish to simplify
\begin{align}
  H_0 \V_2 \ket\psi
  = \sum_O \sum_{\substack{k<\ell\\p<q}} w_{Ok\ell} h_{pq}
  \Pi_{pq} O_{k\ell} \ket\psi,
  \label{eq:H_V2_psi_start}
\end{align}
which has terms with $\set{p,q}=\set{k,\ell}$, terms with
$p,q\notin\set{k,\ell}$, and terms with only one of $p$ or
$q\in\set{k,\ell}$.  The permutation operator $\Pi_{pq}$ acts
trivially on $O_{k_\ell}\ket\psi$ when $\set{p,q}=\set{k,\ell}$ or
$p,q\notin\set{k,\ell}$, leaving terms at each fixed $k,\ell$ of the
form
\begin{align}
  h_{k\ell} + \sum_{\substack{p<q\\p,q\notin\set{k,\ell}}} h_{pq}
  = \sum_{p<q} h_{pq}
  - \sum_{\substack{p\in\set{k,\ell}\\q\notin\set{k,\ell}}} h_{pq}
  = E_0 - h_k - h_\ell + 2 h_{k\ell}
  = E_0 + 2 \bar h_{k\ell},
\end{align}
where
\begin{align}
  \bar h_{k\ell} \equiv h_{k\ell} - \f12\p{h_k + h_\ell}.
\end{align}
The corresponding terms in \eqref{eq:H_V2_psi_start} with
$\set{p,q}=\set{k,\ell}$ or $p,q\notin\set{k,\ell}$ are then
\begin{align}
  E_0 \V_2 \ket\psi
  + \sum_O \sum_{k<\ell} 2 \bar h_{k\ell} w_{Ok\ell} O_{k\ell} \ket\psi.
\end{align}
Using similar techniques, we can simplify the terms in
\eqref{eq:H_V2_psi_start} with only one of $p,q\in\set{k,\ell}$ to
\begin{align}
  \f12 \sum_{k\ne\ell} \sum_q \p{h_{kq} w_{Oq\ell} + w_{Okq} h_{q\ell}}
  O_{k\ell} \ket\psi.
\end{align}
In total,
\begin{align}
  H_0 \V_2 \ket\psi
  &= E_0 \V_2 \ket\psi
  + \sum_{k<\ell} 2\bar h_{k\ell} w_{Ok\ell} O_{k\ell} \ket\psi
  + \f12 \sum_{k\ne\ell} \sum_q \p{h_{kq} w_{Oq\ell} + w_{Okq} h_{q\ell}}
  O_{k\ell} \ket\psi \\
  &= E_0 \V_2 \ket\psi
  + \f12 \sum_{k\ne\ell} \sp{2\bar h_{k\ell} w_{Ok\ell}
    + \sum_q \p{h_{kq} w_{Oq\ell} + w_{Okq} h_{q\ell}}}
  O_{k\ell} \ket\psi.
\end{align}

@@

Altogether, the two-body excitation energies corresponding to
non-trivial eigenstates of $\ut{\check{\m h}}_k$ are
\begin{align}
  \Delta_0 \equiv -2\p{N-1} - 2 h_0 = 0,
  &&
  \Delta_+ \equiv -\varepsilon_+ - 2 h_0 = N,
\end{align}
while the energy of all orthogonal two-body excitations is
\begin{align}
  \Delta_\perp \equiv - 2 h_0 = 2N - 2.
\end{align}

>>>>

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQFS \sum_O \sum_{p<q} w_{Opq} O_{0,1}
  \EQFS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-body
operator $O_{pq}$.  If the two-body operator $O=X\otimes Y$, then
\begin{align}
  H_2^{(1)} \EQFS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} - \col{XY}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V2_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \f12 \sum_O \sum_{p\ne q} \sp{2\bar h_{pq} w_{Opq}
    + \sum_k \p{h_{pk} w_{Okq} + w_{Opk} h_{kq}}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_{pq} - \f12\p{h_p + h_q},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ (indexed by $\Delta$) is an eigenvector
of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w} = \Delta \v{\m w},
  \label{eq:cond_2}
\end{align}
where
\begin{align}
  \check{\m h} \equiv 2\sum_{p<q} \bar h_{pq} \op{\set{p,q}}
  + \sum_{p<q} \ket{\set{p,q}} \sum_{k\notin\set{p,q}}
  \p{h_{pk} \bra{\set{k,q}} + h_{kq} \bra{\set{p,k}}}.
  \label{eq:h_super_mat}
\end{align}
Solving the general eigenvalue problem in \eqref{eq:cond_2} requires
diagonalizing the ${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$
matrix $\check{\m h}$.  When both he interaction Hamiltonian $H_0$
translationally invariant, however, this eigenvalue problem can be
reduced to that of diagonalizing an $\sim N\times N$ matrix; we
discuss this reduction in Appendix \ref{sec:trans_inv}.

@@

We wish to simplify
\begin{align}
  H_0 \V_1 \ket\psi
  = \sum_k \sum_{p<q} h_{pq} v_k \Pi_{pq} X_k \ket\psi,
  \label{eq:H_V1_psi_start}
\end{align}
where without loss of generality we consider a single-body
perturbation $\V_1$ built out of only one single-body operator $X$.
The sum in \eqref{eq:H_V1_psi_start} has terms with $k\in\set{p,q}$,
and terms with $k\notin\set{p,q}$.  In the case of $k\notin\set{p,q}$,
the permutation operator $\Pi_{pq}$ commutes with $X_k$ and
annihilates on $\ket\psi$, leaving terms at each fixed $k$ of the form
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}} h_{pq}
  = \sum_{p<q} h_{pq} - \sum_{\substack{p<q\\k\in\set{p,q}}} h_{pq}
  = E_0 - \f12 \sum_{\substack{p,q\\k\in\set{p,q}}} h_{pq}
  = E_0 - h_k,
\end{align}
where we define $h_{pq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $h_{pq}=h_{qp}$ and $h_{pp}=0$; and
$h_k \equiv \sum_\ell h_{k\ell}$.  The terms in
\eqref{eq:H_V1_psi_start} with $k\notin\set{p,q}$ are then
\begin{align}
  \sum_{\substack{p<q\\k\notin\set{p,q}}}
  h_{pq} v_k \Pi_{pq} X_k \ket\psi
  = \sum_k \p{E_0 - h_k} v_k X_k \ket\psi
  = E_0 \V_1 \ket\psi - \sum_k h_k v_k X_k \ket\psi
\end{align}
while the terms in \eqref{eq:H_V1_psi_start} with $k\in\set{p,q}$ take
the form
\begin{align}
  \sum_{\substack{p<q\\k\in\set{p,q}}}
  h_{pq} v_k \Pi_{pq} X_k \ket\psi
  = \sum_{k<q} h_{kq} v_k X_q \ket\psi
  + \sum_{p<k} h_{pk} v_k X_p \ket\psi
  = \sum_{k,q} h_{kq} v_k X_q \ket\psi
\end{align}
which implies
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \p{\sum_{p,q} h_{pq} v_p X_q - \sum_p h_p v_p X_p}
  \ket\psi.
\end{align}

@@

We wish to simplify
\begin{align}
  H_0 \V_2 \ket\psi
  = \sum_{\substack{k<\ell\\p<q}} w_{k\ell} h_{pq}
  \Pi_{pq} O_{k\ell} \ket\psi,
  \label{eq:H_V2_psi_start}
\end{align}
where without loss of generality we consider a two-body perturbation
$\V_2$ built out of only one two-body operator $O$.  The sum in
\eqref{eq:H_V2_psi_start} has terms with $\set{p,q}=\set{k,\ell}$,
terms with $p,q\notin\set{k,\ell}$, and terms with only one of $p$ or
$q\in\set{k,\ell}$.  The permutation operator $\Pi_{pq}$ acts
trivially on $O_{k\ell}\ket\psi$ when $\set{p,q}=\set{k,\ell}$ or
$p,q\notin\set{k,\ell}$, leaving terms at each fixed $k,\ell$ of the
form
\begin{align}
  h_{k\ell} + \sum_{\substack{p<q\\p,q\notin\set{k,\ell}}} h_{pq}
  = \sum_{p<q} h_{pq}
  - \sum_{\substack{p\in\set{k,\ell}\\q\notin\set{k,\ell}}} h_{pq}
  = E_0 - h_k - h_\ell + 2 h_{k\ell}
  = E_0 + 2 \bar h_{k\ell},
\end{align}
where
\begin{align}
  \bar h_{k\ell} \equiv h_{k\ell} - \f12\p{h_k + h_\ell}.
\end{align}
The corresponding terms in \eqref{eq:H_V2_psi_start} with
$\set{p,q}=\set{k,\ell}$ or $p,q\notin\set{k,\ell}$ are then
\begin{align}
  E_0 \V_2 \ket\psi
  + \sum_{k<\ell} 2 \bar h_{k\ell} w_{k\ell} O_{k\ell} \ket\psi.
\end{align}
Using similar techniques, we can simplify the terms in
\eqref{eq:H_V2_psi_start} with only one of $p,q\in\set{k,\ell}$ to
\begin{align}
  \f12 \sum_{k\ne\ell} \sum_q \p{h_{kq} w_{q\ell} + w_{kq} h_{q\ell}}
  O_{k\ell} \ket\psi.
\end{align}
In total,
\begin{align}
  H_0 \V_2 \ket\psi
  &= E_0 \V_2 \ket\psi
  + \sum_{k<\ell} 2\bar h_{k\ell} w_{k\ell} O_{k\ell} \ket\psi
  + \f12 \sum_{k\ne\ell} \sum_q \p{h_{kq} w_{q\ell} + w_{kq} h_{q\ell}}
  O_{k\ell} \ket\psi \\
  &= E_0 \V_2 \ket\psi
  + \f12 \sum_{k\ne\ell} \sp{2\bar h_{k\ell} w_{k\ell}
    + \sum_q \p{h_{kq} w_{q\ell} + w_{Okq} h_{q\ell}}}
  O_{k\ell} \ket\psi.
\end{align}

@@

Altogether, the excitation energies corresponding to non-trivial
eigenstates of $\ut{\check{\m h}}_k$ are
\begin{align}
  \Delta_0 \equiv -2\p{N-1} - 2 h_0 = 0,
  &&
  \Delta_+ \equiv -\varepsilon_+ - 2 h_0 = N,
\end{align}
while the energy of all orthogonal two-body excitations is
\begin{align}
  \Delta_\perp \equiv - 2 h_0 = 2N - 2.
\end{align}
